dispRity manual

### 2017-06-10 doc/dispRity-manual.Rmd
Manual update

@@

@@

Another way to split the ordinated space (maybe more relevant to palaeobiologists) is to do it according to time.
The `time.subsamples` function allows to create subsamples containing all the elements present at specific points in time or during specific periods.
This functions needs as input an ordinated space and a matching phylogenetic tree.
Two types of time subsamples can be performed by using the `method` option:

 *  discrete time subsamples (or time-binning) using `method = discrete`;

 *  continuous time subsamples (or time-slicing) using `method = continuous`.

@@

 *  defining the boundaries of the time bins (when `method = discrete`);

 *  defining the time slices (when `method = continuous`).

@@

In this example, the taxa were split inside each time-bin according to their age.
Note however that some taxa might span between some time bins and should be included in more than one.
This is possible by providing a table containing the first and last occurrence data:

```{r, eval=FALSE}
## Generating three time bins containing the data present every 40 Ma, including taxa that might
## span between time bins.
(time_bins <- time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree,
    method = "discrete", time = c(120, 80, 40, 0), FADLAD = BeckLee_ages))
```

This generated indeed a `dispRity` object of 3 subsamples.
Note that we can also generate equivalent results by just telling the function that we want three time-bins (subsamples) as follow:

@@

We now have three time bins of 44.50368 million years each.

@@

For the time-slicing method (`method = continuous`), the idea is fairly similar.
This option, however, requires a matrix that contains the ordinated distance between taxa *and* nodes as well as an extra argument being the assumed evolutionary model (via the `model` argument):

 *  `acctran` where the data chosen on each time slice is always the one of the offspring

 *  `deltran` where the data chosen on each time slice is always the one of the descendant

 *  `punctuated` where the data chosen on each time slice is randomly chosen between the offspring or the descendant

 *  `gradual` where the data chosen on each time slice is either the offspring or the descendant depending on branch length

```{r, eval=FALSE}
## Generating four time slices every 40 million years assuming a gradual evolutionary model
(time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
    FADLAD = BeckLee_ages))
## Note that in the same way as before, we can also automatically generate the slices
time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, method = "continuous",
    model = "gradual", time = 4)
```

@@

Once we obtain our different subsamples, it is possible to bootstrap and rarefy them (i.e. pseudo-replicating the data).
The bootstrap will allow us to make each subsamples more robust to outliers and the rarefaction will allow us to compare subsamples with the same number of elements to get rid of eventual sampling problems.
The `boot.matrix` allows to bootstraps and rarefy ordinated matrices in a fast and easy way.
The default options will bootstrap the matrix 1000 times without rarefaction using the "full" bootstrap method (see below).

```{r, eval=FALSE}
boot.matrix(data = BeckLee_mat50)
```

The number of bootstrap pseudo-replicates can be defined using the `bootstraps` option.
The method can be modified by controlling which bootstrap algorithm to use through the `boot.type` argument.
Currently two algorithms are implemented:

 *  `full` where the bootstrapping is entirely stochastic (all the data is bootstrapped)

 *  `single` where only one random elements is replaced by one other random elements each pseudo-replication

This function also allows to rarefy the data using the `rarefaction` argument.
The default argument is `FALSE` but it can be set to `TRUE` to fully rarefy the data (i.e. remove *n* elements for the number of pseudo-replicates, where *n* varies from the maximum number of elements present in each subsamples to a minimum of 3 elements).
It can also be set to one or more `numeric` values to only rarefy to the corresponding number of elements.
One additional important argument is `dimensions` that specifies how many dimensions from the matrix should be used for further analysis.
When left missing, all dimensions from the ordinated matrix are used.

```{r, eval=FALSE}
## Bootstrapping with the single bootstrap method
boot.matrix(BeckLee_mat50, boot.type = "single")

## Bootstrapping with the full rarefaction
boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = TRUE)

## Or with a set number of rarefaction levels
boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = c(6:8,3))

## Removing the last axis 50% last dimensions
boot.matrix(BeckLee_mat50, dimensions = 0.5)

@@

```{r, eval=FALSE}
## Bootstrapping and fully rarefying the crown/stem subsamples
(crown_stemBS <- boot.matrix(crown_stem, bootstraps = 100, rarefaction = TRUE))

## Bootstrapping on the time binning/slicing subsamples
(time_binsBS <- boot.matrix(time_bins, bootstraps = 100))
(time_slicesBS <- boot.matrix(time_slices, bootstraps = 100))
```

@@

## Plotting `dispRity` data

>>>>

<!-- TG: I'm gonna change that to a newest version with the submission -->

@@

This following section contains specific information relative to some functions.
If one of your questions are not covered in these sections, please refer to the function manuals in R or send me an email. <!-- Add anchor to contacts -->

@@

The function `time.subsamples` allows to divide the ordinated matrix into different time subsamples given a dated phylogeny that contains all the elements (i.e. taxa) from the matrix.
Each subsample generated by this function will then contain all the elements present at a specific point in time or during a specific period in time.

Two types of time subsamples can be performed by using the `method` option:

 *  discrete time subsamples (or time-binning) using `method = discrete`;
 *  continuous time subsamples (or time-slicing) using `method = continuous`.

@@

 *  defining the boundaries of the time bins (when `method = discrete`);
 *  defining the time slices (when `method = continuous`).

@@

In this example, the taxa were split inside each time-bin according to their age.
However, the taxa here are considered like single points in time.
It is totally possible that the taxa in question could have a longer longevity (e.g. if the taxonomic focus is at the genera level or above) and that it spans across multiple time bins.
In this case, it is possible to include them in more than one bin by providing a table of First and last occurrence dates (FAD/LAD).
This table should have the taxa names as row names and two columns for respectively the first and last occurrence age:

```{r, eval=FALSE}
## Displaying the table of first and last occurrence dates for each taxa
head(BeckLee_ages)

## Generating time bins including taxa that might span between them
time.subsamples(data = BeckLee_mat50, tree = BeckLee_tree, method = "discrete",
    time = c(120, 80, 40, 0), FADLAD = BeckLee_ages)
```

Note that we can also generate equivalent results by just telling the function that we want three time-bins (subsamples) as follow:

@@

@@

For the time-slicing method (`method = continuous`), the idea is fairly similar.
This option, however, requires a matrix that contains the ordinated distance between taxa *and* nodes as well as an extra argument being the assumed evolutionary model (via the `model` argument).
This model argument is used when the time slice occurs along a branch of the tree rather than on a tip or a node.
The model can be one of the following:

 *  `acctran` where the data chosen along the branch is always the one of the offspring
 *  `deltran` where the data chosen along the branch is always the one of the descendant
 *  `punctuated` where the data chosen along the branch is randomly chosen between the offspring or the descendant
 *  `gradual` where the data chosen along the branch is either the offspring or the descendant depending on branch length

```{r, eval=FALSE}
## Generating four time slices every 40 million years under gradual evolution
time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
    FADLAD = BeckLee_ages)
## Generating four time slices automatically
time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree,
    method = "continuous", model = "gradual", time = 4, FADLAD = BeckLee_ages)
```

@@

One important step in analysing ordinated matrices is to pseudo-replicate the data to seem how robust the results of future analysis are sensitive to outliers in the dataset.
This can be achieved using the function `boot.matrix` to bootstrap or rarefy the ordinated matrix.
The default options will bootstrap the matrix 100 times without rarefaction using the "full" bootstrap method (see below):

```{r, eval=FALSE}
## Default bootstrapping
boot.matrix(data = BeckLee_mat50)
```

The number of bootstrap replicates can be defined using the `bootstraps` option.
The method can be modified by controlling which bootstrap algorithm to use through the `boot.type` argument.
Currently two algorithms are implemented:

 *  `full` where the bootstrapping is entirely stochastic (*n* elements are replaced by any elements drawn from the dat);
 *  `single` where only one random element is replaced by one other random element for each pseudo-replicate.

```{r, eval=FALSE}
## Bootstrapping with the single bootstrap method
boot.matrix(BeckLee_mat50, boot.type = "single")
```

This function also allows to rarefy the data using the `rarefaction` argument.
Rarefaction allows to limit the number of elements to be drawn at each bootstrap replication.
For example, one might be interested in looking at the effect of reducing the number of elements to the results of the analysis.

This can be achieved by using the `rarefaction` option that draws only *n-x* at each bootstrap replicate (where *x* is the number of elements not sampled).
The default argument is `FALSE` but it can be set to `TRUE` to fully rarefy the data (i.e. remove *x* elements for the number of pseudo-replicates, where *x* varies from the maximum number of elements present in each subsamples to a minimum of 3 elements).
It can also be set to one or more `numeric` values to only rarefy to the corresponding number of elements.

```{r, eval=FALSE}
## Bootstrapping with the full rarefaction
boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = TRUE)

## Or with a set number of rarefaction levels
boot.matrix(BeckLee_mat50, bootstraps = 20, rarefaction = c(6:8,3))
```

One additional important argument is `dimensions` that specifies how many dimensions from the matrix should be used for further analysis.
When left missing, all dimensions from the ordinated matrix are used.

```{r, eval = FALSE}
## Removing the last axis 50% last dimensions
boot.matrix(BeckLee_mat50, dimensions = 0.5)

@@

```{r, eval=TRUE}
## Creating subsamples of crown and stem mammals
crown_stem <- custom.subsamples(BeckLee_mat50,
    group = list("crown" = c(16, 19:41, 45:50), "stem" = c(1:15, 17:18, 42:44)))
## Bootstrapping and rarefying these groups
boot.matrix(crown_stem, bootstraps = 200, rarefaction = TRUE)

## Creating time slices subsamples
time_slices <- time.subsamples(data = BeckLee_mat99, tree = BeckLee_tree, 
    method = "continuous", model = "gradual", time = c(120, 80, 40, 0),
    FADLAD = BeckLee_ages)
## Bootstrapping on the time binning/slicing subsamples
boot.matrix(time_slices, bootstraps = 1000)
```

@@

There are many ways for measuring disparity!
In brief, disparity is a summary metric that will represent an aspect of an ordinated space (e.g. a MDS, PCA, PCO, PCoA).
For example, one can look at ellipsoid hyper-volume of the ordinated space
<!-- Cite Donohue -->
, the sum and the product of the ranges and variances
<!-- Cite Wills -->
 or the median position of the elements relative to their centroid.
 <!-- Cite Sive -->
Of course, there are many more examples of metrics one can use for describing some aspect of the ordinated space, with some performing better than other ones at particular descriptive tasks or some being more generalist.

Because of this great diversity of metric, the package `dispRity` has not a unique way to measure disparity but rather proposes to facilitate users to define their own disparity metric that will best suit to their particular analysis.
In fact, the core function of the package, `dispRity`, allows the user to define any metric with the `metric` argument.
However the `metric` argument has to follow certain rules:

1.  it must be composed from one to three `function` objects;
2.  the(se) function(s) must intake as a first argument a `matrix` or a `vector`;
3.  the(se) function(s) must be of one of the three levels described below;
4.  at least one of the functions must be of level 1 or 2 (see below).

### The function levels

The metric functions levels determine the "dimensionality" level of decomposition of the input ordinated matrix.
In other words, each level designate the dimensions of the output: either three (a `matrix`); two (a `vector`); or one (a single `numeric` value) dimensions.

![Illustration of the different levels of output of a function with an input `matrix`](dispRity_fun.png)

#### A level 1 function

A level 1 function will decompose a `matrix` or a `vector` into a single value:

```{r}
## Creating a dummy matrix
dummy_matrix <- matrix(rnorm(12), 4, 3)

## Example of level 1 functions
mean(dummy_matrix)
median(dummy_matrix)
```

Any summary metric such as mean or median are good examples of level 1 functions as they reduce the matrix to a single dimension (i.e. one value).

#### A level 2 function

A level 2 function will decompose a `matrix` into a `vector`.

```{r}
## Defining the function as the product of rows
prod.rows <- function(matrix) apply(matrix, 1, prod)

## A level 2 metric
prod.rows(dummy_matrix)
```

Several level 2 functions are implemented in `dispRity` (see `?dispRity.metric`) such as the `variances` function that calculates the variance of each dimension of the ordinated matrix or the `ranges` one that calculates the range of each dimension.

#### A level 3 function

Finally a level 3 function will transform the matrix into another matrix.
Note that the dimension of the output matrix don't need to match the the input one:

```{r}
## A level 3 metric
var(dummy_matrix)

## A level 3 metric with a forced matrix output
as.matrix(dist(dummy_matrix))
```

### `make.metric`

Of course, functions can be more complex and involve multiple operations such as the `centroids` function (see `?dispRity.metric`) that calculates the euclidean distance between each elements and the a position in the ordinated space.
The `make.metric` function implemented in `dispRity` is designed to help testing and finding the level of the functions.
This function tests:

1.  if your function can deal with a `matrix` or a `vector` as an input;
2.  which is your function's level according to it's output (level 1, 2 or 3, see above);
3.  whether the function can properly be implemented in the `dispRity` function (the function is fed into a `lapply` loop).

For example, let's see if the functions described above are the right levels:

```{r}
## Which level is the mean function? And is able to be used in the dispRity?
make.metric(mean)

## Same questions from the prod.rows function:
make.metric(prod.rows)

## Same questions for the var function:
make.metric(var)
```

A silent version of the function is also available to be included in your code.
This can be done using the option `silent = TRUE` and will simply output the level of the metric.

```{r}
## Testing whether mean is level 1
if(make.metric(mean, silent = TRUE) != "level1") {
    message("The metric is not level 1.")
}
## Testing whether var is level 1
if(make.metric(var, silent = TRUE) != "level1") {
    message("The metric is not level 1.")
}
```

### metric argument in `dispRity`

Using this metric structure, we can easily use any disparity metric in the `dispRity` function as follows:

```{r}
## Measuring disparity as the standard deviation of all the value of the
## ordinated matrix (level 1 function).
summary(dispRity(BeckLee_mat50, metric = sd))

## Measuring disparity as the standard deviation of the variance of each axis of
## the ordinated matrix (level 1 and level 2 functions).
summary(dispRity(BeckLee_mat50, metric = c(sd, variances)))

## Measuring disparity as the standard deviation of the variance of each axis of
## the variance covariance matrix (level 1, level 2 and level 3 functions).

summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)), round = 10)
```

Note that the order of each function in the metric argument does not matter, the `dispRity` function will automatically detect the function levels (using `make.metric`) and apply them to the data in decreasing order (level 3 > level 2 > level 1).

```{r}
## Disparity as the standard deviation of the variance of each axis of the
## variance covariance matrix:
disparity1 <- summary(dispRity(BeckLee_mat50, metric = c(sd, variances, var)), round = 10)

## Same but using a different functions order for the metric argument
disparity2 <- summary(dispRity(BeckLee_mat50, metric = c(variances, sd, var)), round = 10)

## Both ways output the same disparity values:
disparity1 == disparity2
```

#### What level is the disparity?

In these examples, we considered disparity to be a single value.
For example, in the previous one, we defined disparity as the standard deviation of the variances of each column of the variance/covariance matrix (`metric = c(variances, sd, var)`).
This results in single value to be calculated.

In some cases however, it might also be interesting to calculate disparity as a distribution.
For example, one of the metrics commonly used in palaeobiology is the sum of variances of each dimensions (`metric = c(sum, variances)`) [@Wills1994].
However, it might be interesting to look directly at the distribution of the variances itself.

```{r}
## A one dimensional disparity value
(sum_variances1 <- summary(dispRity(BeckLee_mat50, metric = c(median, variances))))

## A two dimensional disparity
(summary(all_variances <- (dispRity(BeckLee_mat50, metric = variances))))
```

Looking at all the variances rather than simply their median can tell us that (by the look of it) the variances of each dimensions seems to be normally distributed.

> It is also possible to reuse an already calculated disparity distribution and summarise it using a level 1 metric:

```{r}
## Transforming the two dimensional one in a one dimensional one
sum_variances2 <- summary(dispRity(all_variances, metric = median))

## Both ways output the same disparity values:
sum_variances1 == sum_variances2
```

### Metrics implemented in `dispRity`

Several disparity metrics are directly implemented in the `dispRity` package.
The detailed list can be found in `?dispRity.metric` along with some explanations on each metric.

Level | Name | Description | Source | 
------|------|---------------------------------------------------|--------|
3 | `ellipse.volume`<sup>1</sup> | The volume of the ellipsoid of the space | [@DonohueDim] |
3 | `convhull.surface` | The surface of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
3 | `convhull.volume` | The volume of the convex hull formed by all the elements | [`geometry`](https://cran.r-project.org/web/packages/geometry/index.html)`::convhulln` |
3 | `hypervolume` | The volume of the ordinated space | [`hypervolume`](https://cran.r-project.org/web/packages/hypervolume/index.html)`::hypervolume` |
3 | `diagonal` | The longest distance in the ordinated space (like the diagonal in two dimensions) | `dispRity` |
2 | `ranges` | The range of each dimension | `dispRity` |
2 | `variances` | The variance of each dimension | `dispRity` |
2 | `centroids`<sup>2</sup> | The distance between each element and the centroid of the ordinated space | `dispRity` |
1 | `mode.val` | The modal value | `dispRity` |

1: The functions estimation of the eigen value only works for MDS or PCoA ordinations (not PCA).

2: Note that by default, the centroid is the centroid of the elements.
It can however be fixed to an arbitrary value by using the `centroid` argument `centroids(space, centroid = rep(0, ncol(space)))`, for example the origin of the ordinated space.

### Equations and implementations:
Some of the functions described below are directly implemented in the `dispRity` package and do not require any dependencies  ([see implementation here](https://github.com/TGuillerme/dispRity/blob/master/R/dispRity.metric.R)).

\begin{equation}
    ellipse.volume = \frac{\pi^{k/2}}{\Gamma(\frac{k}{2}+1)}\displaystyle\prod_{i=1}^{k} (\lambda_{i}^{0.5})
\end{equation}
With *k* being the number of dimensions, $\lambda_i$ is the eigen value of each dimension.

\begin{equation}
    diagonal = \sqrt{\sum_{i=1}^{k}|max(k_i) - min(k_i)|}
\end{equation}
With *k* being the dimensions.

\begin{equation}
    ranges = |max(k_i) - min(k_i)|
\end{equation}
With *k* being the dimensions.

\begin{equation}
    variances = \sigma^{2}{k_i}
\end{equation}
With *k* being the dimensions and $\sigma^{2}$ their variance.

\begin{equation}
    centroids = \sqrt{\sum_{i=1}^{n}{({k}_{n}-Centroid_{k})^2}}
\end{equation}
With *n* being each element in the ordinated space, *k* being its dimensions and $Centroid_{k}$ their mean (or can be set to an arbitrary value) and.

### Metrics run through
Here is a run through of the main metrics implemented in `dispRity`.
First, we will create a dummy ordinated space using the `space.maker` utility function:

```{r}
## Creating a 10*5 normal space
set.seed(1)
dummy_space <- space.maker(10, 5, rnorm)
```

#### Volumes and surface metrics
The functions `ellipse.volume`, `convhull.surface`, `convhull.volume` and `hyper.volume` all measure the surface or the volume of the ordinated space occupancy:

```{r}
## Calculating the ellipsoid volume
summary(dispRity(dummy_space, metric = ellipse.volume))
```
> Because there is only one subsamples (i.e. one matrix) in the dispRity object, this operation is the equivalent of doing simply `ellipse.volume(dummy_space)` (with rounding).

```{r}
## Calculating the convex hull surface
summary(dispRity(dummy_space, metric = convhull.surface))

## Calculating the convex hull volume
summary(dispRity(dummy_space, metric = convhull.volume))
```

The convex hull functions make a (good) estimation of the multidimensional properties of the ordinated space.
For the full and correct calculation of the volume of the ordinated space, it is possible to us the `hyper.volume` function that intakes more options but takes longer to calculate.

```{r}
## Calculating the true multidimensional volume
summary(dispRity(dummy_space, metric = hyper.volume))

## Calculating the volume with different options to be passed to hypervolume::hypervolume
summary(dispRity(dummy_space, metric = hyper.volume, repsperpoint = 100, quantile = 0.5))
```
> Cautionary note: measuring volumes in a high number of dimensions can be strongly affected by the [Curse of dimensionality](https://en.wikipedia.org/wiki/Curse_of_dimensionality) which often results in near 0 disparity values.

#### Ranges, variances and diagonal

The functions `ranges`, `variances` and `diagonal` all measure properties of the ordinated space base on it's dimensions properties (they are also less affected by the Curse of dimensionality):

`ranges` and `variances` both work on the same principle and measure the range/variance of each dimensions:

```{r}
## Calculating the ranges of each dimensions in the ordinated space
ranges(dummy_space)

## Calculating disparity as the distribution of these ranges
summary(dispRity(dummy_space, metric = ranges))

## Calculating disparity as the sum and the product of these ranges
summary(dispRity(dummy_space, metric = c(sum, ranges)))
summary(dispRity(dummy_space, metric = c(prod, ranges)))

## Calculating the variances of each dimensions in the ordinated space
variances(dummy_space)

## Calculating disparity as the distribution of these variances
summary(dispRity(dummy_space, metric = variances))

## Calculating disparity as the sum and the product of these variances
summary(dispRity(dummy_space, metric = c(sum, variances)))
summary(dispRity(dummy_space, metric = c(prod, variances)))
```

The `diagonal` function measures the multidimensional diagonal of the whole space (i.e. in our case the longest euclidean distance in our 5 dimensional space):

```{r}
## Calculating the ordinated space's diagonal
summary(dispRity(dummy_space, metric = diagonal))
```

> This metric is only mathematically valid if the dimensions within the space are all orthogonal!

#### Centroids metrics

The `centroids` metric allows to measure the position of the different elements compared to a fixed point in the ordinated space.
By default, this function measures the distance between each elements and the central point between them:

```{r}
## Calculating the distribution of the distances between each elements and the centroid
summary(dispRity(dummy_space, metric = centroids))

## Calculating disparity as the median value of these distances
summary(dispRity(dummy_space, metric = c(median, centroids)))
```

It is however possible to fix the coordinates of a specific point in the ordinated space as long as it has the correct number of dimensions:

```{r}
## Calculating the distance between the elements and the origin of the ordinated space
summary(dispRity(dummy_space, metric = centroids, centroid = c(0,0,0,0,0)))

## Calculating disparity as the distance between the elements and a specific point
summary(dispRity(dummy_space, metric = centroids, centroid = c(0,1,2,3,4)))
```
